Linting Linearity in Core/System FC

### 2023-01-11 report.tex
Stuck on linear type background cohesion

@@

\begin{itemize}
    \item Linear types were only added to the language roughly \emph{31 years
        after} Haskell's inception. It is an especially difficult endeavour to
        add linear types to a well-established language with a large library
        ecosystem and many active users, rather than to develop the language
        from the ground up with linear types in mind, and the successful
        approach as implemented in GHC 9.0, the leading Haskell compiler, was
        based on Linear Haskell\cite{linearhaskell}, where a linear type system
        designed with retaining backwards-compatibility and allowing code reuse
        across linear and non-linear users of the same libraries in mind was
        described.
    \item Linear types permeate Haskell down to (its) \textbf{Core}, the intermediate
        language Haskell is desugared to. \textbf{Core} is a very small
        explicitly typed language to which multiple Core-to-Core optimizing
        transformations are applied in succession. The final Core program can be
        typechecked very fast (because of the explicit type annotations) and
        doing so serves as a sanity check to the correction of the
        source transformations. If the resulting optimized Core program doesn't
        typecheck, something \emph{very wrong} has happened in a transformation!
\end{itemize}

Even though Linear Haskell was successful in integrating linear types in an
existing language 
% In spite of the smooth integration of linear types in an existing language
with regard to backwards compatibility and (re)usability,

The advent of linear types in Haskell bring forth

@@

Programming languages with linear type systems might differ in their definition of linear
resources, but usually \emph{variables} are considered linear resources
% That is, we can rely on the type system to prove properties regarding the usage of certain resources.
The \emph{Hello World} of linearly typed programs is a linearly typed pointer.
In a non-memory managed language, an allocated pointer must be freed
\emph{exactly once}.

\begin{code}
let p = malloc(5)
    p' = store(p, "x")
    (p'', v) = load(p)
    free p''
\end{code}

@@

Haskell is a functional programming language defined by the "Haskell
Report"\cite{jones1999haskell,marlow2010haskell} and whose \emph{de-facto}
implementation is GHC, the Glasgow Haskell Compiler\cite{GHC}. Haskell is a
\emph{non-strict}, \emph{purely functional} language, i.e., functions cannot
have side effects or mutate data, and contrary to many programming languages,
arguments are \emph{not} evaluated when passed to functions, but rather are only
evaluated when the value is needed.

@@

% TODO:
Is it outside the scope of this report to discuss the merits and disadvantages
of laziness and purely functional properties? Não estou nada contente com esta
secção em geral. Acho que preciso de escrever o resto para conseguir alguma
coerência.

@@

\subsection{Linear Haskell}

@@

\subsection{Core Haskell}

>>>>

% TODO: linear functions to allow safe/controlled use of reified language
% implementation objects like *single-use continuations* (single use
% continuations? can we make them faster?)

@@

\begin{itemize}
    \item Linear types were only added to the language roughly \emph{31 years
        after} Haskell's inception. It is an especially difficult endeavour to
        add linear types to a well-established language with a large library
        ecosystem and many active users, rather than to develop the language
        from the ground up with linear types in mind, and the successful
        approach as implemented in GHC 9.0, the leading Haskell compiler, was
        based on Linear Haskell\cite{linearhaskell}, where a linear type system
        designed with retaining backwards-compatibility and allowing code reuse
        across linear and non-linear users of the same libraries in mind was
        described. We describe Linear Haskell in detail in
        section~\ref{linear-haskell-section}.
    \item Linear types permeate Haskell down to (its) \textbf{Core}, the intermediate
        language Haskell is desugared to. \textbf{Core} is a very small
        explicitly typed language to which multiple Core-to-Core optimizing
        transformations are applied in succession. The final Core program can be
        typechecked very fast (because of the explicit type annotations) and
        doing so serves as a sanity check to the correction of the
        source transformations. If the resulting optimized Core program doesn't
        typecheck, something \emph{very wrong} has happened in a transformation!
        We discuss Core in detail in section~\ref{core-section}.
\end{itemize}

In theory, the linearity annotations in \textbf{Core} should allow for the same
typechecking verification we had with the existing type annotations prior to the
addition of linear types, that is, our Core program with linearity annotations
should be typechecked after the optimizing transformations...

% Even though Linear Haskell was successful in integrating linear types in an
% existing language 
% In spite of the smooth integration of linear types in an existing language
% with regard to backwards compatibility and (re)usability,

% The advent of linear types in Haskell bring forth

@@

A linear type system is a type system in which ...!

% A linear type system is a type system in which statements such as "this function
% will use its argument exactly once" can be formally expressed and checked.

Programming languages with linear type systems might differ in their definition of linear
resources, but usually \emph{variables} are considered linear resources?
% That is, we can rely on the type system to prove properties regarding the usage of certain resources.
% The \emph{Hello World} of linearly typed programs is a linearly typed pointer.
% In a non-memory managed language, an allocated pointer must be freed
% \emph{exactly once}.

% \begin{code}
% let p = malloc(5)
%     p' = store(p, "x")
%     (p'', v) = load(p)
%     free p''
% \end{code}

@@

Haskell is a functional programming language defined by the "Haskell
Report"\cite{jones1999haskell,marlow2010haskell} and whose \emph{de-facto}
implementation is GHC, the Glasgow Haskell Compiler\cite{GHC}. Haskell is a
\emph{lazy}, \emph{purely functional} language, i.e., functions cannot
have side effects or mutate data, and contrary to many programming languages,
arguments are \emph{not} evaluated when passed to functions, but rather are only
evaluated when the value is needed.

% TODO: File example

@@

% TODO:
% Is it outside the scope of this report to discuss the merits and disadvantages
% of laziness and purely functional properties?

@@

\subsection{Linear Haskell\label{linear-haskell-section}}

@@

\subsection{Core Haskell\label{core-section}}
